The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-05-15 papers/pair-correlation/paper.tex

\section{Previous approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  These works differ in what density is
used when evaluating the radial distribution function $g(r,\eta)$.

Several recent works~\cite{gloor2007prediction, gross2009density} have
assumed a mean-density approximation for the pair distribution
function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.

>>>>

\section{Previous approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  Work on these approximations date back to
Toxvaerd \cite{toxvaerd1973statistical} and have been tested
rigorously for both functional and non-functional based theories
\cite{kalos1977structure,osborn1980monotonic,mccoy1981comparison}.
The approximations differ in what density is used when evaluating the
radial distribution function $g(r,\eta)$ or in how radial
distributions evaluated at different densities are combined to yield
an effective function \cite{toxvaerd1973statistical}.

One common approximation \cite{gloor2007prediction, gross2009density}
assumes a mean-density approximation for the pair distribution
function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and
Methfessel~\cite{fischer1980born,harris1987comment} introduce the
approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is an integral of the density over a spherical volume that
is later defined as one of the fundamental measures in Fundamental
Measure Theory~\cite{rosenfeld1989free}:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.
